---
title: "Biostat 203B Homework 4"
subtitle: Due Mar 18 @ 11:59PM
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

```{r, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

Display machine information:
```{r}
sessionInfo()
```
Load database libraries and the tidyverse frontend:
```{r}
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(miceRanger))
```

## Q1. Missing data

Through the Shiny app developed in HW3, we observe abundant missing values in the MIMIC-IV ICU cohort we created. In this question, we use multiple imputation to obtain a data set without missing values.

0. Read following tutorials on the R package miceRanger for imputation: <https://github.com/farrellday/miceRanger>, <https://cran.r-project.org/web/packages/miceRanger/vignettes/miceAlgorithm.html>.

    A more thorough book treatment of the practical imputation strategies is the book [*_Flexible Imputation of Missing Data_*](https://stefvanbuuren.name/fimd/) by Stef van Buuren. 

1. Explain the jargon MCAR, MAR, and MNAR.

**Solution**

**MCAR**
**If the probability of being missing is the same for all cases, then the data** 
**are said to be missing completely at random (MCAR), which implies that causes** 
**of the missing data are unrelated to the data.**

**MAR**
**If the probability of being missing is the same only within groups defined** 
**by the observed data, then the data are missing at random (MAR).** 

**MNAR**
**If neither MCAR nor MAR holds, then we speak of missing not at random (MNAR).**
**MNAR means that the probability of being missing varies for reasons that are** 
**unknown to us.**

2. Explain in a couple of sentences how the Multiple Imputation by Chained Equations (MICE) work.

**Solution**

**Firstly MICE checks the raw data sets, and finds and replaces those missing**
**data with some random values, where those random values are from other**
**observations of this variable. Then using the specified method of prediction,**
**MICE will find the closest values for all missing values and fill them up**
**variable by variable (column by column). According to the number of**
**iterations we specify, MICE will repeat above steps and return a final**
**imputation.**

3. Perform a data quality check of the ICU stays data. Discard variables with substantial missingness, say >5000 `NA`s. Replace apparent data entry errors by `NA`s.

**Solution**

**Step 1: filter the substantial missingness.

```{r}
icucohort_tble = read_rds("/Users/a123456/biostat-203b-2022-winter/hw3/mimiciv_shiny/icu_cohort.rds")
Datacheck = c(1 : 59)
for (i in 1 : ncol(icucohort_tble)) {
  Datacheck[i] = sum(is.na(icucohort_tble[ , i]))
}
Datacheck
for (i in c(1 : 59)) {
  if(Datacheck[i] > 5000){
    print(colnames(icucohort_tble[1, i]))
  }
}
```

**From the result of data quality check, we will discard the variables**
**"deathtime", "edregtime", "edouttime", "dod", "hadm_to_death".**

**Step 2: detect and replace apparent data entry errors.

```{r}
icucohort_tble1 = icucohort_tble
icucohort_tble = select(icucohort_tble, -c("deathtime", "edregtime", 
                                           "edouttime", "dod", "hadm_to_death"))
Datacheck = Datacheck[Datacheck <= 5000]
```

```{r}
dataerror = list()
for(i in c(8 : 53)) {
  colpoint = icucohort_tble[ , i]
  colname = colnames(colpoint)
  if(!is.timepoint(colpoint[[1]])){
    if(!is.character(colpoint[[1]])){
      dataerror[[i-7]] = boxplot.stats(colpoint[[1]])$out
#      ggplot() +
#  geom_boxplot(mapping = aes(x = colname, y = colpoint[[1]]))
    }
  }
}

```

4. Impute missing values by `miceRanger` (request $m=3$ data sets). This step is computational intensive. Make sure to save the imputation results as a file. Hint: Setting `max.depth=10` in the `miceRanger` function may cut some computing time.

5. Make imputation diagnostic plots and explain what they mean.

6. Choose one of the imputed data sets to be used in Q2. This is **not** a good idea to use just one imputed data set or to average multiple imputed data sets. Explain in a couple of sentences what the correct Multiple Imputation strategy is.

## Q2. Predicting 30-day mortality

Develop at least two analytic approaches for predicting the 30-day mortality of patients admitted to ICU using demographic information (gender, age, marital status, ethnicity), first lab measurements during ICU stay, and first vital measurements during ICU stay. For example, you can use (1) logistic regression (`glm()` function in base R or keras), (2) logistic regression with lasso penalty (glmnet or keras package), (3) random forest (randomForest package), or (4) neural network (keras package).

1. Partition data into 80% training set and 20% test set. Stratify partitioning according the 30-day mortality status.

2. Train the models using the training set.

3. Compare model prediction performance on the test set.

