---
title: "Biostat 203B Homework 3"
subtitle: Due Feb 25 @ 11:59PM
author: YANLONG BAI
output:
  # ioslides_presentation: default
  html_document:
    toc: true
    toc_depth: 4
---

Display machine information:
```{r}
sessionInfo()
```

Load database libraries and the tidyverse frontend:
```{r}
suppressPackageStartupMessages(library(dbplyr))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(lubridate))
library(DBI)
library(bigrquery)
```

## Q1. Compile the ICU cohort in HW2 from the Google BigQuery database 

Below is an outline of steps.

1. Load the service account token.
```{r}
# path to the service account token 
if (Sys.info()[["nodename"]] == "biostat-203b-teaching-server") {
  # on teaching server
  satoken <- "/mnt/mimiciv/1.0/biostat-203b-2022winter-3fdc2392ac39.json"
} else {
  # on my own Mac laptop
  satoken <- 
    "/Users/a123456/biostat-203b-2022-winter/biostat-203b-2022winter-3fdc2392ac39.json"
}
# BigQuery authentication using service account
bq_auth(
  path = satoken,
  # email = "mimiciv-bigquery@biostat-203b-2022winter.iam.gserviceaccount.com",
  # scopes = c("https://www.googleapis.com/auth/bigquery",
  #   "https://www.googleapis.com/auth/cloud-platform")
)
```



1. Connect to BigQuery database `mimic4_v1_0_203b` in GCP (Google Cloud Platform), using the billing account `biostat-203b-2022winter`.
```{r}
# Connect to the BigQuery database `biostat-203b-2022winter.mimic4_v1_0_203b`
con <- dbConnect(
    bigrquery::bigquery(),
    project = "biostat-203b-2022winter",
    dataset = "mimic4_v1_0_203b",
    billing = "biostat-203b-2022winter"
)
con
```

2. List all tables in the `mimic4_v1_0_203b` database.
```{r}
dbListTables(con)
```

3. Connect to the `icustays` table.
```{r}
# full ICU stays table
icustays_tble <- tbl(con, "icustays") %>%
  show_query() %>%
  print(width = Inf)
```
We only keep the first ICU stay. Following code is kind of a hack, using the `summarise_all(min)` function. It seems that `slice_min()`, `slice_head()`, `distinct(, .keep_all = TRUE)` don't work with `dbplyr`+`bigrquery` at the moment.
```{r}
icustays_tble <- icustays_tble %>% 
  select(subject_id, intime) %>% 
  group_by(subject_id) %>% 
  summarise_all(min) %>%
  left_join(icustays_tble, by = c("subject_id", "intime")) %>%
  show_query() %>%
  print(width = Inf)
```


4. Connect to the `admissions` table and only keep the patients who have a match in `icustays_tble` (according to `subject_id` and `hadm_id`).

**Solution**

```{r}
admissions_tble <- tbl(con, "admissions") %>%
  semi_join(select(icustays_tble, subject_id, hadm_id), 
            by = c("subject_id", "hadm_id")) %>%
  show_query() %>%
  print(width = Inf)
summarise(admissions_tble, n = n())
```

5. Connect to the `patients` table and only keep the patients who have a match in `icustays_tble` (according to `subject_id`).

**Solution**

```{r}
patients_tble <- tbl(con, "patients") %>%
  semi_join(select(icustays_tble, subject_id), by = c("subject_id")) %>%
  show_query() %>%
  print(width = Inf)
summarise(admissions_tble, n = n())
```

6. Connect to the `labevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the lab items listed in HW2. 

**Solution**

```{r}
interest_lab_mes = c(50912, 50971, 50983, 50902, 50882, 51221, 51301,
                     50931, 50960, 50893)
lab_mes = tbl(con, "d_labitems") %>%
  filter(itemid %in% interest_lab_mes) %>%
  select(-fluid, -category, -loinc_code)
labevents_tble = tbl(con, "labevents") %>%
  inner_join(lab_mes, by = "itemid") %>%
  inner_join(select(icustays_tble, subject_id), by = "subject_id") %>%
  show_query() %>%
  print(width = Inf)
summarise(labevents_tble, n = n())
```

Only keep the first lab measurements during ICU stay and pivot lab items to become variables/columns.

**Solution**

```{r}
labevents_tble = left_join(labevents_tble, 
                           select(icustays_tble, subject_id, intime, outtime),
                           by = "subject_id") %>%
  filter(charttime < outtime &
         charttime > intime) %>%
  select(subject_id, itemid, charttime, valuenum)
  
labcharttm = select(labevents_tble, subject_id, itemid, charttime) %>%
  group_by(subject_id, itemid) %>%
  summarise_all(min) %>%
  ungroup()

labevents_tble = inner_join(labevents_tble, labcharttm, 
                            by = c("subject_id", "itemid", "charttime")) %>%
  pivot_wider(names_prefix = "", names_from = c("itemid"), 
              values_from = c("charttime","valuenum")) %>%
  print(width = Inf)
summarise(labevents_tble, n = n())
rm(lab_mes)
rm(interest_lab_mes)
rm(labcharttm)
gc()
```

7. Connect to `chartevents` table and retrieve a subset that only contain subjects who appear in `icustays_tble` and the chart events listed in HW2.

**Solution**

```{r}
interest_vit = c(220045, 220181, 220179, 223761, 220210)
vit = tbl(con, "d_items") %>%
  filter(itemid %in% interest_vit)
chartevents_tble = tbl(con, "chartevents") %>%
  inner_join(select(vit, itemid), by = "itemid") %>%
  inner_join(select(icustays_tble, subject_id), by = "subject_id") %>%
  select(subject_id, stay_id, charttime, itemid, valuenum) %>%
  show_query() %>%
  print(width = Inf)
summarise(chartevents_tble, n = n())
```

Only keep the first chart events during ICU stay and pivot chart events to become variables/columns.

```{r}
chartevents_tble = left_join(chartevents_tble, 
                             select(icustays_tble, subject_id, intime, outtime), 
                             by = "subject_id") %>%
  filter(charttime < outtime &
         charttime > intime) %>%
  select(subject_id, itemid, charttime, valuenum)
  
chartevecharttm = select(chartevents_tble, subject_id, itemid, charttime) %>%
  group_by(subject_id, itemid) %>%
  summarise_all(min) %>%
  ungroup()

chartevents_tble = inner_join(chartevents_tble, chartevecharttm, 
                              by = c("subject_id", "itemid", "charttime")) %>%
  pivot_wider(names_prefix = "", names_from = c("itemid"), 
              values_from = c("charttime","valuenum")) %>%
  print(width = Inf)
summarise(chartevents_tble, n = n())

rm(chartevecharttm)
rm(vit)
rm(interest_vit)
gc()

```

8. Put things together. This step is similar to Q7 of HW2. Using one chain of pipes `%>%` to perform following data wrangling steps: (i) start with the `icustays_tble` for the first ICU stay of each unique patient, (ii) merge in admissions and patients tables, (iii) keep adults only (age at admission >= 18), (iv) merge in the labevents and chartevents tables, (v) create an indicator for 30-day mortality, (vi) save the final tibble to an `icu_cohort.rds` R data file in the `mimiciv_shiny` folder.

```{r}
# make a directory mimiciv_shiny
if (!dir.exists("mimiciv_shiny")) {
  dir.create("mimiciv_shiny")
}
```

```{r}
mimic_icu_cohort = left_join(icustays_tble, admissions_tble, 
                             by = c("subject_id", "hadm_id")) %>%
  left_join(patients_tble, by = "subject_id") %>%
  mutate(age_hadm = anchor_age + year(admittime) - anchor_year) %>%
  filter(age_hadm >= 18) %>%
  left_join(labevents_tble, by = "subject_id") %>%
  left_join(chartevents_tble, by = "subject_id") %>%
  mutate(hadm_to_death = if(is.na(dod)) Inf 
         else as.character(dod-admittime)) %>%
  mutate(thirty_day_mort = if(str_length(hadm_to_death) != 8){
    if(as.numeric(substr(hadm_to_death, 5, 6)) <= 30) 1 else 0} else 0) %>%
  print(width = Inf)  %>%
  collect() %>%
  write_rds(
    "/Users/a123456/biostat-203b-2022-winter/hw3/mimiciv_shiny/icu_cohort.rds")
```

Close database connection and clear workspace.
```{r, eval = F}
dbDisconnect(con)
rm(list = ls())
```

## Q2. Shiny app

Develop a Shiny app for exploring the ICU cohort data created in Q1. The app should reside in the `mimiciv_shiny` folder. The app should provide easy access to the graphical and numerical summaries of variables (demographics, lab measurements, vitals) in the ICU cohort.

